### 2025-10-03 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
2025NSF: draft of proposed work

A core insight that is guiding this enterprise is that geometric structure in both classical and quantum mechanics are equivalent to entropic structures. In classical mechanics, a single degree of freedom is a two dimensional symplectic manifold. In this case, symplectic form $\omega$ is a volume form and therefore specifying $\omega$ is equivalent to specifying the Liouville measure $\mu(U) = \int_U \omega$. In classical statistical mechanics, the entropy of a statistical ensemble is given by $S(\rho) = - \int \rho \log \rho d\mu$ where $\rho$ is the probability density associated to the ensemble. For a uniform distribution with support $U$, the entropy reduce to $S(\rho_U) = \log \mu(U)$. Since the logarithm is an invertible function, one can recover the measure from the entropy. That is, specifying the entropy associated to all uniform distributions allows us to reconstruct the Liouville measure, and therefore the symplectic form. In the general case, Darboux coordinates allow, at least locally, to factorize the space into multiple independent degrees of freedom, where the volume is the product of the corresponding areas. The corresponding entropic property is that the entropy for the joint distribution is the sum of the marginals if they are independently distributed. Therefore, if the distribution of all possible marginals are specified, the symplectic form can be recovered.

In quantum mechanics, the geometry is given by the Born rule
\begin{equation}
	p(\phi|\psi) = \frac{\< \phi | \psi \>\< \psi | \phi \>}{\< \psi | \psi \>\< \phi | \phi \>}.
\end{equation}
Given two states $\psi$ and $\phi$, the ensemble $\rho_{\psi\phi}$ given by their equal mixture has entropy $S(\rho_{\psi\phi}) = I\left(\frac{1 + \sqrt{p(\phi,\psi)}}{2},\frac{1 - \sqrt{p(\phi,\psi)}}{2} \right)$ where $I(x,y) = -x \log x -y \log y$ is the Shannon entropy. The entropy as a function of $p(\psi,\phi)$ is invertible. Therefore, given the entropy of the equal mixtures of all pairs of states we can reconstruct the Born rule and, therefore, the state space.

While the states spaces of classical (i.e. symplectic manifolds) and quantum mechanics (i.e. projective Hilbert spaces) are rather different, the space of ensembles of both have a similar structure. They are both convex subsets of a topological vector space, equipped with a strictly convex entropy function. The core idea, then, is to use ensembles as a way to generalize the notion of physical states, and the entropy to characterize the geometry in a way that is independent of the particulars of each physical theory. In the past year we have done significant preliminary work that shows that not only this approach is possible, but is extremely productive. The proposed work is to turn those preliminary insight in an axiomatic specification that is fully physically motivated. Such an approach, if successful, would at least provide the following benefits:
\begin{enumerate}
	\item clarify exactly what physical premises guarantee the basic geometric structures of physical theories, allowing a more principled formulation of current physical theories and search for new ones;
	\item provide a generalized structure for information theory, possibly generalizing results from both classical and quantum information theory
	\item provide useful constrains to more precisely formulate physical theories in the infinite dimensional case.
\end{enumerate}

@@

In a nutshell, the goal is to find a very limited set of axioms that can be guaranteed to be valid in any physical theory, and that allow us to generalize a good number of results from all physical theories. The notion of statistical ensemble (i.e. the collection of all outputs of a preparation procedure) is particularly suitable for generalization for a number of both practical and conceptual considerations. Our preliminary work shows that there are three, possible four, distinct set of requirements for an ensemble space $\Ens$, the set of ensemble given by a physical theory.

@@

The second requirement is the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This is insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs). The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to physically untenable cases). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space. Unfortunately, we haven't found such a structure already studied in the literature. While it would be extremely useful for us to study these spaces on their own merit, we choose to concentrate on problems where our expertise in the mathematical foundations of physical theories is more critical, as in the next requirement.

The third requirement is characterizing the variability of elements within a statistical ensemble. Though the instance within ensembles are never formalized, the variability within an ensemble is physically significant as it will influence, for example, the variability of statistical measurements. As we showed in previous work, entropy quantifies the variability of the elements within a distribution, and which gives a physically meaningful starting point. The proposed work is to find the exact specification of the axiom to see whether it can also guarantee that ensemble spaces must be metric spaces.

The hardest conceptual problem is that entropy is defined differently in each space. In a classical discrete space, the entropy is $-\sum p_i \log p_i$; in classical statistical mechanics is $- \int \rho \log \rho d\mu$ where $\mu$ is the Liouville measure; in quantum mechanics is $-\tr(\rho \log \rho)$. How can we give a single definition for these very different expressions? Note that the expressions are different precisely because they are in terms of the specific representation of ensembles in specific theory. Our approach is to simply characterize the entropy by its properties with respect to the topology and the convex structure. In our preliminary properties, for example, the entropy $S : \Ens \to \mathbb{R}$ has the following properties:
\begin{itemize}
	\item Continuity
	\item Strict concavity: $S(p\ens[a] + \bar{p} \ens[b]) \geq p S(\ens[a]) + \bar{p} S(\ens[b])$ with the equality holding if and only if $\ens[a] = \ens[b]$
	\item Upper variability bound: there exists a universal function $I(p_1, p_2)$ (i.e. the same for all ensemble spaces) such that $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p, \bar{p}) + p S(\ens[a]) + \bar{p} S(\ens[b])$; if the equality holds, $\ens[a]$ and $\ens[b]$ are orthogonal, noted $\ens[a] \ortho \ens[b]$
	\item Mixtures preserve orthogonality: $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$
\end{itemize}
These properties are justified by the physics. For example, if we are performing a statistical mixture between two ensemble, the average variability cannot decrease, which justifies the concavity. These relatively simple properties allow us, for example, to show that $I(p,\bar{p})$, the maximum entropy increase during mixing, must be $ - \kappa \left(p \log p + \bar{p} \log \bar{p}\right)$, the Shannon entropy. The proof is general, meaning that if we are right that those axioms must hold for any physical theory, then the Shannon entropy is of truly general applicability. Moreover, in the case of both classical and quantum mechanics, states can be decomposed into orthogonal components, which means the upper variability is valid with an equal. That allows to recover the expression for the entropy in the three different cases.

Our preliminary work shows that an entropy satisfying those axioms forces the convex space to be cancellative, meaning that it embeds into a vector space. That is, if we pick an internal point $\ens \in \Ens$, we can define a vector space $V_{\ens}$ such that $\Ens \subset V_{\ens}$. If we pick to different internal points $\ens \in \Ens$, we can show that there is an invertible affine map between $V_{\ens[a]}$ is $V_{\ens[b]}$. That is, the ensemble space embeds in an affine space. We can also show that it is ``directionally bounded,'' meaning its intersection with any affine line is finite. It is an open question whether the axioms are enough to prove that the embedding is continuous in a topological vector space, and whether the set is bounded.

We can also show that the entropy can be used to define geometric structures directly in the ensemble space. For example, we can define the mixing entropy as $MS(\ens[a], \ens[b]) = S\left(\frac{1}{2}\ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2} S(\ens[a]) + \frac{1}{2} S(\ens[b])\right)$. If we express the entropy in bits, the mixing entropy satisfies the following:
\begin{enumerate}
	\item non-negativity: $MS(\ens[a], \ens[b]) \geq 0$
	\item identity of indiscernibles: $MS(\ens[a], \ens[b]) = 0 \iff \ens[a]=\ens[b]$
	\item unit boundedness: $MS(\ens[a], \ens[b]) \leq 1$
	\item maximality of orthogonals: $MS(\ens[a], \ens[b]) = 1 \iff \ens[a] \ortho \ens[b]$
	\item symmetry: $MS(\ens[a], \ens[b]) = MS(\ens[b], \ens[a])$
\end{enumerate}
It is almost a distance function, as it does not satisfy the triangle inequality, but it recovers the Jensen-Shannon divergence (JSD) in both classical and quantum mechanics. The JSD is the square of a distance function in both classical and quantum information theory, though our current axioms do not guarantee $MS$ to be so. The $MS$ can be used to create open balls, and show that the topology must be Hausdorff. The current axioms do not guarantee the open balls to be convex.

If we assume we are in finite dimensions and that the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite which allows us to define a metric tensor. Let $V_{\ens[a]}$ be the embedding vector space with $\ens[a] \in \Ens$ as origin. This will be a differentiable manifold. On a chart whose coordinates are linear with respect to the affine structure we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessian are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor. In classical and quantum mechanics it recovers the Fisher information metric. Note that, if were to show continuous embedding into a locally convex topological vector space, this construction can be extended to the infinite dimensional case.

understand whether the existence of a physically meaningful entropy function is enough to guarantee that the space of statistical ensembles of any physical theory is a convex subset of a locally convex topological vector space equipped with a metric. That is, we should not simply ``declare'' it to be so, but make sure that the mathematical starting points are justified by physical arguments, so that we can

>>>>

A core insight that is guiding this enterprise is that geometric structures in both classical and quantum mechanics are equivalent to entropic structures. In classical mechanics, a single degree of freedom is a two dimensional symplectic manifold. In this case, symplectic form $\omega$ is a volume form and therefore specifying $\omega$ is equivalent to specifying the Liouville measure $\mu(U) = \int_U \omega$. In classical statistical mechanics, the entropy of a statistical ensemble is given by $S(\rho) = - \int \rho \log \rho d\mu$ where $\rho$ is the probability density associated to the ensemble. For a uniform distribution with support $U$, the entropy reduce to $S(\rho_U) = \log \mu(U)$. Since the logarithm is an invertible function, one can recover the measure from the entropy. That is, specifying the entropy associated to all uniform distributions allows us to reconstruct the Liouville measure, and therefore the symplectic form. In the general case, Darboux coordinates allow, at least locally, to factorize the space into multiple independent degrees of freedom, where the volume is the product of the corresponding areas. The corresponding entropic property is that the entropy for the joint distribution is the sum of the marginals if the independent degrees of freedom are independently distributed. Therefore, if the distribution of all possible marginals are specified, the symplectic form can be recovered.

In quantum mechanics, the geometry is given by the Born rule
\begin{equation}
	p(\phi|\psi) = \frac{\< \phi | \psi \>\< \psi | \phi \>}{\< \psi | \psi \>\< \phi | \phi \>}.
\end{equation}
Given two states $\psi$ and $\phi$, the ensemble given by their equal mixture, mathematically $\rho_{\psi\phi} = \frac{1}{2}|\psi\>\<\psi|+\frac{1}{2}|\phi\>\<\phi|$,  has entropy $S(\rho_{\psi\phi}) = I\left(\frac{1 + \sqrt{p(\phi,\psi)}}{2},\frac{1 - \sqrt{p(\phi,\psi)}}{2} \right)$ where $I(x,y) = -x \log x -y \log y$ is the Shannon entropy. Therefore the entropy is an invertible function of $p(\psi,\phi)$. This means that, given the entropy of the equal mixtures of all pairs of states we can reconstruct the Born rule and, therefore, the state space.

While the states spaces of classical (i.e. symplectic manifolds) and quantum mechanics (i.e. projective Hilbert spaces) are rather different, the space of ensembles of both have a similar structure. They are both convex subsets of a topological vector space, equipped with a strictly convex entropy function. The core idea, then, is to use ensembles as a way to generalize the notion of physical states, and the entropy to characterize the geometry in a way that is independent of the particulars of each physical theory. In the past year we have done significant preliminary work that shows that not only this approach is possible, but is extremely productive. The proposed work is to turn those preliminary insight in an axiomatic specification that is fully physically motivated. If successful, such an approach will provide the following benefits:
\begin{enumerate}
	\item clarify exactly what physical premises guarantee the basic geometric structures of physical theories, allowing a more principled formulation of current physical theories as well as the search for new ones;
	\item provide a generalized structure for information theory, possibly allowing future generalization of results from both classical and quantum information theory
	\item provide useful constrains to more precisely formulate physical theories in the infinite dimensional case.
\end{enumerate}

@@

The goal is to find a very limited set of axioms to define the entropic structure of a physical theory, which can be justified from physical considerations, show that it imposes a geometric structure on a generic space of statistical ensembles and that it recovers the standard geometric structures of classical and quantum mechanics. Furthermore, the same structure should be connected to topological and measure theoretic structures used in physics.

%and Other that are required by any physical theory to define its entropic structure, which  and construct a general connection to geometry, calculus and topology, thus giving a generalized precise mathematical characterization of physical theories. In our numerous conversations with mathematicians, physicists and philosophers, it is clear that such work is sorely needed, and, since it lies at the intersection of different fields, it requires an interdisciplinary approach.

We want to stress that finding the correct starting point is not just a mathematical issue, but also a physics problem. Let us explore this problem through an example. The space $M_1([0,1])$ of probability measures defined over the $[0,1]$ interval is an infinite dimensional vector space. As Aliprantis and Border note: ``Since there is more than one topology of interest on an infinite dimensional space, the choice of topology is a key modeling decision that can have economic as well as technical consequences.'' Unbeknowst to many physicists, this is true for physics as well. If we choose a weak topology, we are saying the convergence of expectation values of all continuous functions is enough to determine convergence of ensembles. While this may seem plausible at first glance, we can find a weakly convergence sequence for which the entropy does not converge. If the entropic structure of a physical theory is really equivalent to its geometric structure, not only the entropy should be continuous, but it is likely that the topology should be the one generated by the entropic structure. Ultimately, answering these questions means understanding both the physical and mathematical consequences of those decisions.

Our preliminary work convinced us that the notion of statistical ensemble (i.e. the collection of all outputs of a preparation procedure) seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the math, the structure of statistical ensembles of both classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form ``whenever I prepare a system according to this procedure, I will find this outcome.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that a physical theory must \emph{at least} posit a space of ensembles. Our preliminary work shows that there are three, possibly four, distinct set of requirements for an ensemble space $\Ens$, the set of ensemble given by a physical theory.

@@

The second requirement is the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This is insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs). The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to physically untenable cases). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space. While we haven't found such a structure already studied in the literature, and it would extremely useful to us to see what can be generalized from topological vector spaces to topological convex spaces, this is a fairly well defined mathematical question at this point, and we believe that our expertise in the mathematical foundations of physical theories is better spent in continuing finding other well defined starting points.

The third requirement, central to the work, is characterizing the variability of elements within a statistical ensemble. Though we imagine ensembles to be a collection of instances, these instances are never formalized as they need to be left unspecified to leave the theory general. However, the variability of the elements within an ensemble is physically significant as it will influence, for example, the variability of statistical measurements. As we showed in previous work, entropy quantifies the variability of the elements within a distribution, and which gives us a physically meaningful starting point. The primary goal of the proposed work, then, is to find what axiomatic properties must this variability indicator have, and to understand what are their conceptual, physical and mathematical implications.

The hardest conceptual problem we found is that the entropy is defined differently in each space. In a classical discrete space, the entropy is $-\sum p_i \log p_i$; in classical statistical mechanics is $- \int \rho \log \rho d\mu$ where $\mu$ is the Liouville measure; in quantum mechanics is $-\tr(\rho \log \rho)$. How can we give a single definition for these very different expressions? Note that the expressions are different precisely because they are in terms of the specific representation of ensembles in specific theory. Our approach is to simply characterize the entropy by its properties with respect to the topology and the convex structure. In our preliminary work, for example, the entropy $S : \Ens \to \mathbb{R}$ has the following properties:
\begin{itemize}
	\item Continuity
	\item Strict concavity: $S(p\ens[a] + \bar{p} \ens[b]) \geq p S(\ens[a]) + \bar{p} S(\ens[b])$ with the equality holding if and only if $\ens[a] = \ens[b]$
	\item Upper variability bound: there exists a universal function $I(p_1, p_2)$ (i.e. the same for all ensemble spaces) such that $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p, \bar{p}) + p S(\ens[a]) + \bar{p} S(\ens[b])$; if the equality holds, $\ens[a]$ and $\ens[b]$ are orthogonal, noted $\ens[a] \ortho \ens[b]$
	\item Mixtures preserve orthogonality: $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$
\end{itemize}
These properties are justified by the physics. For example, if we are performing a statistical mixture between two ensemble, the average variability cannot decrease, which justifies the concavity. These relatively simple properties allow us, for example, to show that $I(p,\bar{p})$, the maximum entropy increase during mixing, must be $ - \kappa \left(p \log p + \bar{p} \log \bar{p}\right)$, the Shannon entropy. The proof is general, meaning that if we are right that those axioms must hold for any physical theory, then the Shannon entropy is of truly general applicability. Moreover, in the case of both classical and quantum mechanics, states can be decomposed into orthogonal components, which means the upper variability is valid with an equal, and can be used to calculate the entropy from that decomposition. That allows to recover the expression for the entropy in the three different cases.

Our preliminary work shows that an entropy satisfying those axioms forces the convex space to be cancellative, meaning that it embeds into a vector space. That is, if we pick an internal point $\ens \in \Ens$, we can define a vector space $V_{\ens}$ such that there exists an embedding $\iota : \Ens \hookrightarrow  V_{\ens}$ such that $\iota(p \ens[a] + \bar{p} \ens[b]) = p \iota(\ens[a]) + \bar{p} \iota(\ens[b])$ and $\iota(\ens) = 0$. If we pick two different internal points $\ens[a],\ens[b] \in \Ens$, we can show that there is an invertible affine map between $V_{\ens[a]}$ is $V_{\ens[b]}$. That is, the ensemble space embeds in an affine space. We can also show that it is ``directionally bounded,'' meaning its intersection with any affine line is finite. It is an open question whether the axioms are enough to prove that the embedding is continuous in a topological vector space, and whether the set is bounded.

We can also show that the entropy can be used to define geometric structures directly on the ensemble space. For example, we can define the mixing entropy as $MS(\ens[a], \ens[b]) = S\left(\frac{1}{2}\ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2} S(\ens[a]) + \frac{1}{2} S(\ens[b])\right)$. If we express the entropy in bits, the mixing entropy satisfies the following:
\begin{enumerate}
	\item non-negativity: $MS(\ens[a], \ens[b]) \geq 0$
	\item identity of indiscernibles: $MS(\ens[a], \ens[b]) = 0 \iff \ens[a]=\ens[b]$
	\item unit boundedness: $MS(\ens[a], \ens[b]) \leq 1$
	\item maximality of orthogonals: $MS(\ens[a], \ens[b]) = 1 \iff \ens[a] \ortho \ens[b]$
	\item symmetry: $MS(\ens[a], \ens[b]) = MS(\ens[b], \ens[a])$
\end{enumerate}
It is almost a distance function, as it does not satisfy the triangle inequality, but it recovers the Jensen-Shannon divergence (JSD) in both classical and quantum mechanics. The JSD is the square of a distance function in both classical and quantum information theory, though our current axioms do not guarantee $MS$ to be so. It is an open question whether enough physical constraints exist for that to be the case. The $MS$ can be used to create open balls, and show that the topology must be Hausdorff. An interesting conjecture is whether an Hausdorff second countable topological convex space is metrizable, given that this is true for topological vector spaces. The current axioms do not guarantee the open balls to be convex. If we were to find sufficient physically motivated conditions that imply that the entropic open balls are convex and generate the topology, as we suspect it should be, the topology would be locally convex and therefore the entropic structure also gives us enough structure to define calculus.

If we assume we are in finite dimensions and that the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite which allows us to define a metric tensor. Let $V_{\ens[a]}$ be the embedding vector space with $\ens[a] \in \Ens$ as origin. This will be a differentiable manifold. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessian are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is an Hessian manifold. In classical and quantum mechanics it recovers the Fisher information metric. The ultimate goal would be to show that, in general case, the embedding is continuous into a locally convex topological vector space so that this construction can be extended to the infinite dimensional case. This would essentially show that the entropy provides, in general, the geometric structure of the state space of any physical theory.

The fourth structure, which is left for future work, is the generalization of the Lie brackets given by the Poisson brackets in the classical case and the commutators in the quantum case. These describe the relationship between the generators of transformations that leave the geometric structure (i.e. symplectic form and inner product respectively) unchanged, which means, as outlined before, is equivalent to the geometric structure. That is, it describes the group of deterministic and reversible motions, those that leave the entropy unchanged. Given the tight connection between the entropic structure and the Lie algebraic structure, it is unclear whether this will be recovered automatically once the entropic structure is correctly axiomatized, or whether it will require additional axioms. For this reason, we postpone the study of this aspect.

Having given an overview of the specifics of the problems, the proposed work consists of first finalizing and publishing some of the preliminary results, which include the equivalence of geometric and entropic structures in physical theories, the ability to recover the Shannon entropy in the generalized case and the three different entropies in the discrete/continuous/quantum cases. At the same time, explore whether the axiomatic can be expanded, without sacrificing potentially interesting physical cases, to recover potentially useful properties, such as the metrizability and local convexity of the topology generated by the entropy.
